Compositional Verification of Concurrent C Programs with Search Structure Templates

### 2023-05-02 concurrency/CONCUR 2023/paper.tex
modified postcondition of traverse

%maybe inline this instead of making it a figure?
\begin{figure}[ht]
	\centering
	\begin{mathpar}
		% spec of inRange
		{\color{blue}
			\forall \  m. \left\langle \texttt{pn} \mapsto (\texttt{p, n}) \ \ast \ 
			\nodeboxrep(\texttt{n})  \ \ast \ \mathsf{R}(\texttt{n}) \ \big| \ \treerep\ m \
			\right\rangle
		}
		\\ 
		\texttt{traverse(pn, k)} 
		\\
		{\color{blue}
			\left\langle \mathit{res}. \ \exists \  \texttt{n'}.
			\begin{array}{l} \texttt{pn} \mapsto (\texttt{n', n'}) \ \ast 
				\ (\mathit{res} \leftrightarrow (\texttt{n'->t} = \texttt{NULL}))  \ \ast \ \\ \nodeboxrep(\texttt{n'}) \ast \ \mathsf{R}(\texttt{n'})\ \ast \ \texttt{k} \in \mathsf{range}(\texttt{n'})
			\end{array}
			\ \Bigg| \ \treerep\ m \
			\right\rangle
		}
	\end{mathpar}
	\caption{Specification of \texttt{traverse} for the lock-coupling template\wm{Not sure how we want to write this, but when $\mathit{res}$ is false, \lstinline{k} should be the key of \lstinline{n'}, right?}}
	\label{fig:traverse_lock}
\end{figure}

@@

Recall that in the lock-coupling pattern, each thread always holds at least one lock while traversing the data structure, ensuring that other threads cannot interfere with the traversal or invalidate the ongoing search. Consequently, before calling the \texttt{traverse} function, a thread must acquire the lock for the current node $\texttt{n}$ and obtain the resources $\mathsf{R}(\texttt{n})$ guarded by the lock. These resources may depend on the specific data structure, but always include a piece of \emph{ghost state} describing the current state of the node (its key, value, and key range) shared between the invariant $\mathsf{R}$ and the abstract state $\treerep$, ensuring that the lock and the abstract state agree on the contents of the node. Formally, the lock invariant is defined by 
\begin{align*}
&\mathsf{node\_contents}(\texttt{n}, \mathsf{k}, \mathsf{v}, \mathsf{range}) \triangleq \mathsf{ghost\_node(\texttt{n}, k, v, range)} \ \ast \ \mathsf{node\_data}(\texttt{n}, \mathsf{k}, \mathsf{v})\\
&\mathsf{R}(\texttt{n}) \triangleq \exists\ \mathsf{k}, \mathsf{v}, \mathsf{range}.\ \mathsf{node\_contents}(\texttt{n}, \mathsf{k}, \mathsf{v}, \mathsf{range})\end{align*}
where the definition of $\mathsf{node\_data}$ is supplied by the target data structure.

@@

The local precondition of \lstinline{traverse} includes both the node handle $\nodeboxrep(\texttt{n})$ and its contents $\mathsf{R}(\texttt{n})$. The output of \lstinline{traverse} is a new node \lstinline{n'} in the data structure such that the key \texttt{k} falls within the range of \lstinline{n'}. The local postcondition then includes the handle of the new node $\nodeboxrep(\texttt{n'})$, its contents $\mathsf{R}(\texttt{n'})$, and a Boolean variable $\mathit{res}$ indicating whether \texttt{n'->t} is \texttt{NULL} or not. The variable $\mathit{res}$ is $\mathsf{true}$ if $\texttt{flag}$ returns 1, and $\mathsf{false}$ if 0, mirroring its Boolean state. The key to the correctness of the \texttt{traverse} function is the loop invariant for the top-level loop, which expresses that in each iteration, \lstinline{traverse} holds the lock on a node that has \lstinline{k} in its range:
\begin{align*}\mathsf{traverse\_inv}(\texttt{pn}) \triangleq\ \exists \ \texttt{p}, \texttt{n'}, \mathsf{k'}, \mathsf{v'}, \mathsf{range'}.\ &\texttt{pn} \mapsto (\texttt{p, n'}) \ast \texttt{k} \in \mathsf{range'}\ \ast \  \nodeboxrep(\texttt{n'})  \ \ast \\  &\mathsf{node\_contents}(\texttt{n'}, \mathsf{k'}, \mathsf{v'}, \mathsf{range'})
\end{align*}

\begin{figure}[!ht]
	$\color{blue}
	\forall \  m. \left\langle \texttt{pn} \mapsto (\texttt{p, n}) \ \ast \ 
	\nodeboxrep(\texttt{n})  \ \ast \ \mathsf{R}(\texttt{n}) \ \big| \ \treerep\ m \
	\right\rangle$
		\lstinputlisting[language=C, style=myStyle, mathescape=true]{proof_lock_traverse.c}
	$\color{blue}
	\left\langle \mathit{res}. \ \exists \  \texttt{n'}.
	\begin{array}{l} \texttt{pn} \mapsto (\texttt{n', n'}) \ \ast 
		\ (\mathit{res} \leftrightarrow (\texttt{n'->t} = \texttt{NULL}))  \ \ast \ \\ \nodeboxrep(\texttt{n'}) \ast \ \mathsf{R}(\texttt{n'})\ \ast \ \texttt{k} \in \mathsf{range}(\texttt{n'})
	\end{array}
	\ \Bigg| \ \treerep\ m \
	\right\rangle$
	\caption{Proof outline of the lock-coupling \texttt{traverse} function}
	\label{proof_lock_traverse}
\end{figure}

@@

\begin{figure}[h]
	\centering
	\begin{mathpar}
		% spec of inRange
		{\color{blue}
			\forall \  m. \left\langle
			\begin{array}{c}
				\texttt{pn} \mapsto (\texttt{p, n}) \ \ast \ \nodeboxrep(\texttt{n}) \ \big| \ \treerep\ m
			\end{array}
			\right\rangle
		}
		\\ 
		\texttt{traverse(pn, k)} 
		\\
		{\color{blue}
			\left\langle \mathit{res}. \ \exists \ \texttt{n'}.
			\begin{array}{l}
				\texttt{pn} \mapsto (\texttt{n', n'}) \ \ast \ (\mathit{res} \leftrightarrow (\texttt{n'->t} = \texttt{NULL}))  \ \ast \ 
				\\ \nodeboxrep(\texttt{n'}) \ \ast \ \mathsf{R}(\texttt{n'}) \ \ast \ \texttt{k} \in \mathsf{range}(\texttt{n'})
			\end{array}
			\ \Bigg| \ \treerep\ m \
			\right\rangle
		}
	\end{mathpar}
	\caption{Specification of \texttt{traverse} for the give-up template}
	\label{fig:traverse_giveup}
\end{figure}

>>>>

%maybe inline this instead of making it a figure?
\begin{figure}[ht]
	\centering
	\begin{mathpar}
		% spec of inRange
		{\color{blue}
			\forall \  m. \left\langle \texttt{pn} \mapsto (\texttt{p, n}) \ \ast \ 
			\nodeboxrep(\texttt{n})  \ \ast \ \mathsf{R}(\texttt{n}) \ \big| \ \treerep\ m \
			\right\rangle
		}
		\\ 
		\texttt{traverse(pn, k)} 
		\\
		{\color{blue}
			\left\langle \mathit{res}. \ \exists \  \texttt{n'}.
			\begin{array}{l} \texttt{pn} \mapsto (\texttt{n', n'}) \ \ast \nodeboxrep(\texttt{n'}) \ \ast \ 
				\\ 
				\mathsf{if} \ \mathit{res} \ \mathsf{then} \ \mathsf{node\_contents}(\texttt{n'}, \cdot, \mathsf{range}) \ 
				\\ \ \ \ \ \ \ \ \ \ \mathsf{else} \ \mathsf{node\_contents}(\texttt{n'}, (\mathsf{k}, \mathsf{v'}), \mathsf{range})
			\end{array}
			\ \Bigg| \ \treerep\ m \
			\right\rangle
		}
	\end{mathpar}
	\caption{Specification of \texttt{traverse} for the lock-coupling template%\wm{Not sure how we want to write this, but when $\mathit{res}$ is false, \lstinline{k} should be the key of \lstinline{n'}, right?}
	}
	\label{fig:traverse_lock}
\end{figure}

@@

Recall that in the lock-coupling pattern, each thread always holds at least one lock while traversing the data structure, ensuring that other threads cannot interfere with the traversal or invalidate the ongoing search. Consequently, before calling the \texttt{traverse} function, a thread must acquire the lock for the current node $\texttt{n}$ and obtain the resources $\mathsf{R}(\texttt{n})$ guarded by the lock. These resources may depend on the specific data structure, but always include a piece of \emph{ghost state} describing the current state of the node (its key, value, and key range) shared between the invariant $\mathsf{R}$ and the abstract state $\treerep$, ensuring that the lock and the abstract state agree on the contents of the node. Formally, the lock invariant is defined by 
\begin{align*}
	&\mathsf{node\_contents}(\texttt{n}, (\mathsf{k}, \mathsf{v}), \mathsf{range}) \triangleq \mathsf{ghost\_node(\texttt{n}, (\mathsf{k}, \mathsf{v}), range)} \ \ast \ \mathsf{node\_data}(\texttt{n}, \mathsf{k}, \mathsf{v})\\
	&\mathsf{R}(\texttt{n}) \triangleq \exists\ \mathsf{k}, \mathsf{v}, \mathsf{range}.\ \mathsf{node\_contents}(\texttt{n}, (\mathsf{k}, \mathsf{v}), \mathsf{range})\end{align*}
where the definition of $\mathsf{node\_data}$ is supplied by the target data structure.

@@

The local precondition of \lstinline{traverse} includes both the node handle $\nodeboxrep(\texttt{n})$ and its contents $\mathsf{R}(\texttt{n})$. The output of \lstinline{traverse} is a new node \lstinline{n'} in the data structure such that the key \texttt{k} falls within the range of \lstinline{n'}. The local postcondition then includes the handle of the new node $\nodeboxrep(\texttt{n'})$, its contents $\mathsf{R}(\texttt{n'})$, and a Boolean variable $\mathit{res}$ indicating whether \texttt{n'->t} is \texttt{NULL} or not. The variable $\mathit{res}$ is $\mathsf{true}$ if $\texttt{flag}$ returns 1, and $\mathsf{false}$ if 0, mirroring its Boolean state. The key to the correctness of the \texttt{traverse} function is the loop invariant for the top-level loop, which expresses that in each iteration, \lstinline{traverse} holds the lock on a node that has \lstinline{k} in its range:
\begin{align*}\mathsf{traverse\_inv}(\texttt{pn}) \triangleq\ \exists \ \texttt{p}, \texttt{n'}, \mathsf{k'}, \mathsf{v'}, \mathsf{range'}.\ &\texttt{pn} \mapsto (\texttt{p, n'}) \ast \texttt{k} \in \mathsf{range'}\ \ast \  \nodeboxrep(\texttt{n'})  \ \ast \\  &\mathsf{node\_contents}(\texttt{n'}, (\mathsf{k'}, \mathsf{v'}), \mathsf{range'})
\end{align*}

\begin{figure}[!ht]
	$\color{blue}
	\forall \  m. \left\langle \texttt{pn} \mapsto (\texttt{p, n}) \ \ast \ 
	\nodeboxrep(\texttt{n})  \ \ast \ \mathsf{R}(\texttt{n}) \ \big| \ \treerep\ m \
	\right\rangle$
		\lstinputlisting[language=C, style=myStyle, mathescape=true]{proof_lock_traverse.c}
	$\color{blue}
	\left\langle \mathit{res}. \ \exists \  \texttt{n'}.
	\begin{array}{l} \texttt{pn} \mapsto (\texttt{n', n'}) \ \ast \nodeboxrep(\texttt{n'}) \ \ast \ 
		\\ 
		\mathsf{if} \ \mathit{res} \ \mathsf{then} \ \mathsf{node\_contents}(\texttt{n'}, \cdot, \mathsf{range}) \ 
		\\ \ \ \ \ \ \ \ \ \ \mathsf{else} \ \mathsf{node\_contents}(\texttt{n'}, (\mathsf{k}, \mathsf{v'}), \mathsf{range})
	\end{array}
	\ \Bigg| \ \treerep\ m \
	\right\rangle$
	\caption{Proof outline of the lock-coupling \texttt{traverse} function}
	\label{proof_lock_traverse}
\end{figure}

@@

\begin{figure}[h]
	\centering
	\begin{mathpar}
		% spec of inRange
		{\color{blue}
			\forall \  m. \left\langle
			\begin{array}{c}
				\texttt{pn} \mapsto (\texttt{p, n}) \ \ast \ \nodeboxrep(\texttt{n}) \ \big| \ \treerep\ m
			\end{array}
			\right\rangle
		}
		\\ 
		\texttt{traverse(pn, k)} 
		\\
		{\color{blue}
			\left\langle \mathit{res}. \ \exists \  \texttt{n'}.
			\begin{array}{l} \texttt{pn} \mapsto (\texttt{n', n'}) \ \ast \nodeboxrep(\texttt{n'}) \ \ast \ 
				\\ 
				\mathsf{if} \ \mathit{res} \ \mathsf{then} \ \mathsf{node\_contents}(\texttt{n'}, \cdot, \mathsf{range}) \ 
				\\ \ \ \ \ \ \ \ \ \ \mathsf{else} \ \mathsf{node\_contents}(\texttt{n'}, (\mathsf{k}, \mathsf{v'}), \mathsf{range})
			\end{array}
			\ \Bigg| \ \treerep\ m \
			\right\rangle
		}
	\end{mathpar}
	\caption{Specification of \texttt{traverse} for the give-up template}
	\label{fig:traverse_giveup}
\end{figure}
